Propositions as Sessions, Mechanically

### 2015-04-14 paper/formalisation.tex
Polished CPS term translation section.

The translation of GV terms to CP processes is given, following
\citeauthor{Wadler:2014}, using continuation-passing style. In this style a
well-typed term in GV is parameterised by a name representing the channel
which communicates the result of the translation (a CP process). Due to the
locally nameless representation, the translation of GV terms required a
procedure to infer the type of terms in order to specify the binding type for
the corresponding CP process. So, the translation takes the environment, in
which the term is typed, as an argument to query the type of free
variables. Unfortunately, this query returns a type of \coqe!option typ!
(\coqe!option! is equivalent to \tm{Maybe} in Haskell) to account for it
possibly not appearing in the environment, but in fact since only well-typed
GV terms are translated it is guaranteed to return something! Therefore, the
definition is unnecessarily cluttered with impossible error cases.

The term translation was initially designed using a monadic approach to
streamline handling of the possible error cases in retreiving names from the
environment. The coq-ext-lib
library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
definitions for monad, functor and applicative type classes and associated
instances for basic Coq types. It is similar in style to Haskell's support for
these objects. Unfortunately, this definition did not yield a suitable
function to prove the translation well-typed because the function needs to be
able to operate structurally on not only terms, but also the environment and
the processes being constructed. An example, translating choice, is shown in
Figure~\ref{fig:trans}. The translation of case considers \coqe!NL! and
\coqe!NR! in a separate environment to \coqe!M! and types the resulting case
term in the combined environment. Additionally, the subterms need to be
considered as opened terms using cofinite quantification in order to yield a
suitable induction principle.

An alternative formulation, which avoids the need to handle error cases, is to
first define only the positive cases using an inductive type, the so-called
\textit{graph relation}, and then define a dependent function that only
considers the positive cases for which the relation is
defined~\cite{McKinna:2009}. However, several issues were encountered when
attempting to specify the translation. The graph relation requires much of the
clutter present in both the \coqe!cp_rule! and \coqe!wt_tm! definitions such
as linear enviornments and cofinite quantification.  Consider again
Figure~\ref{fig:trans}. unfortunately this definition is wrong because the
process P is not being opened and therefore during the proof we are unable to
reason about the typability of P opened with a sufficiently fresh variable the
reason is that P is a subprocess and therefore is not locally closed note that
I am using \citeauthor{Wadler:2014}'s style but using an LNCQ notation for
clarity. In order to construct the cut and output processes, the type of
\tm{M} and \tm{N} must be computed. the proof of cps translation using the
graph relation is a work in progress. the main difficulty is that the
representation requires opening both GV terms and CP processes in hypotheses
which would appear to duplicate a lot of the overhead of their respective rule
definitions.  again, representation is important even if the proof has still
eluded us.  while it is remains to be seen whether or not the graph relation
assists us in these matters it is nonetheless worth trying to express a
property in multiple ways

Roughly, the theorm aiming to prove is:
\begin{coq}
Theorem cps_trans_wt:
  forall $\Phi$ M T z P $\Gamma$ $\GD$
         (WT: $\Phi$ $\tpgv$ M$\of{T}$)
         (NIN: z $\notin$ GVFV M $\cup$ dom $\Phi$)
         (ENV: trans_env $\Phi$ = $\GD$)
         (PER: Permutation $\Gamma$ ($\GD$ ++ z $\sim$ $\sem{T}$t))
         (CP: transTm M $\Phi$ z P),
    P $\tpcp$ $\Gamma$.
\end{coq}

>>>>

The translation of GV terms to CP processes is given, following
\citeauthor{Wadler:2014}, using continuation-passing style. In this style a
well-typed term in GV is parameterised by a name representing the channel
which communicates the result of its translation (a CP process). The aim is to
prove the following theorem about the translation:
\begin{coq}
Theorem cps_trans_wt:
  forall $\Phi$ M T z P $\Gamma$ $\GD$
         (WT: $\Phi$ $\tpgv$ M$\of{T}$)
         (NIN: z $\notin$ GVFV M $\cup$ dom $\Phi$)
         (ENV: trans_env $\Phi$ = $\GD$)
         (PER: Permutation $\Gamma$ ($\GD$ ++ z $\sim$ $\sem{T}$t))
         (CP: transTm M $\Phi$ z P),
    P $\tpcp$ $\Gamma$.
\end{coq}
where \coqe!z! is the channel parameterising the translation and
\coqe!transTm! is described below. The theorem states that if a GV term
\coqe!M! is well-typed then its translation will be a well-typed process in CP
with some accompanying translations on environments.

Due to the locally nameless representation, the translation of GV terms
requires a procedure to infer the type of terms in order to specify binder(s)
in the corresponding CP process (see the right-hand side of
Figure~\ref{fig:trans}). So the translation takes the environment, in which
the term is typed, as an argument to query the type of free
variables. Unfortunately, this query returns a type of \coqe!option typ!
(\coqe!option! is equivalent to \tm{Maybe} in Haskell) to account for the name
possibly not appearing in the environment, but in fact since only well-typed
GV terms are translated it is guaranteed to return something! Therefore, the
definition is unnecessarily cluttered with impossible error cases.

The term translation was initially designed using a monadic approach to
streamline handling of the possible error cases. The coq-ext-lib
library~\footnote{https://github.com/coq-ext-lib/coq-ext-lib} provides
definitions for monad, functor and applicative type classes and associated
instances for basic Coq types. It is similar in style to Haskell's support for
these objects. Unfortunately, this definition did not yield a suitable
function to prove the translation well-typed because the function needs to be
able to operate structurally on not only terms, but also the environment and
the processes being constructed.

An alternative formulation, which avoids the need to handle error cases, is to
first define only the positive cases using an inductive type, the so-called
\textit{graph relation}, and then define a dependent function that only
considers the positive cases for which the relation is
defined~\cite{McKinna:2009}. An example, translating choice, is shown in
Figure~\ref{fig:trans}. The translation of choice considers \coqe!NL! and
\coqe!NR! in a separate environment to \coqe!M! and types the resulting case
term in the combined environment. The \coqe!NL! and \coqe!NR! subterms need to
be considered as opened terms using cofinite quantification in order to yield
suitable induction hypotheses. The graph relation is able to specify these
constraints in contrast to the monadic function definition. However, several
issues were encountered when attempting to specify the translation. The graph
relation requires much of the clutter present in both the \coqe!cp_rule! and
\coqe!wt_tm! definitions such as linear enviornments and cofinite
quantification, and unfortunately the specification of choice in
Figure~\ref{fig:trans} is not quite correct. The process \coqe!P! is not being
opened in the \key{TR} assumption yet in the conclusion it is not locally
closed. It would appear that, in addition to opening GV subterms, the graph
relation must also open CP subprocesses even in cases where there was no
binder in the original GV term e.g. function
application~\cite{Wadler:2014}. Attempts to resolve this issue are left for
future work as is the proof of \coqe!cps_trans_wt!, but it appears the graph
relation approach is the most promising. These difficulties again illustrate
the importance of the representation choice.
